---
title: "UKBB Genetic Ancestry Prediction"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Intersect the reference with target databases

Extract the reference panel's SNP set from each UKBB imputed chromosome, merge
chromosomes, and write allele frequencies for the intersected UKBB genotypes.

```{bash ukbb-intersect}
plink=/path/to/plink_v1.9/plink
plink2=/path/to/plink2
cohort=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc
refname=$(echo $cohort | sed "s/.*\///g");
snps=${cohort}.snplist
path=~/midProject/data/cohorts/ukbb/hg37/

## (1) Extract snps from UKBB

for i in {1..22}; do
file=/path/to/ukbiobank/imputation/bgen_to_pgen/ukb_imp_chr${i}_v3;
output=${path}intersected_plink/ukb_imp_chr${i}_v3;
$plink2 \
--pfile ${file} \
--snps-only 'just-acgt' \
--set-all-var-ids chr@:#:\$r:\$a \
--extract $snps --keep-allele-order \
--make-bed --out $output; done

# Concat chromosomes
> ${path}mergelist.txt
for i in {1..22}; do
echo ${path}intersected_plink/ukb_imp_chr${i}_v3 >> ${path}mergelist.txt;
done

$plink \
--merge-list ${path}mergelist.txt --keep-allele-order --make-bed --out ${path}ukb_imp_v3_${refname}

### output AF
$plink2 \
--bfile ${path}ukb_imp_v3_${refname} \
--freq \
--out ${path}ukb_imp_v3_${refname}
```

## Get variant outliers

Compare allele frequencies between the reference and the UKBB-intersected variant
set, and flag variants with large frequency differences or rare MAF as outliers
to exclude (Supplementary Fig. 10).

```{r ukbb-variant-outliers}
library(ggplot2)
training="almarri_hgdp1kgpqc_hg37"
pcs=6
test="ukb_imp_v3"

#plot AF
af_ref <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", training, "_", pcs, "pcs.afreq"), header = T, sep = "\t")
af_projection <- read.csv(paste0("~/midProject/data/cohorts/ukbb/hg37/ukb_imp_v3_", training, ".afreq"), header = T, sep = "\t")
af_ref <- af_ref[af_ref$ID %in% af_projection$ID, ]
outlier_af <- data.frame(ID = af_projection[af_projection$ALT_FREQS < 0.05 | af_projection$ALT_FREQS > 0.95 | af_ref$ALT_FREQS - af_projection$ALT_FREQS > 0.4 | af_projection$ALT_FREQS - af_ref$ALT_FREQS > 0.4, ]$ID)

png(paste0("~/midProject/results/figures/af_", test, "X", training, pcs,"pc.pdf"), width = 300, height = 300)
plot(af_ref$ALT_FREQS, af_projection$ALT_FREQS,
     pch = 19,
     col = rgb(0.5, 0.5, 0.5, 0.3),
     xlab = "Almarri+HGDP+1KGP AF", 
     ylab = "UKBB AF")
points(
  af_ref[af_ref$ID %in% outlier_af$ID, ]$ALT_FREQS,
  af_projection[af_projection$ID %in% outlier_af$ID, ]$ALT_FREQS,
  col = rgb(1, 0, 0, 0.6), pch = 19
)

dev.off()

#save outliers
write.csv(outlier_af, file = paste0("~/midProject/data/cohorts/ukbb/hg37/", test, "_", training, "af_outliers.tsv"), col.names=F, row.names = F, quote = F)
```

## Exclude outliers

Remove the flagged variant outliers from the reference panel before running PCA.

```{bash exclude-outliers}
# exclude ukbb variant outliers from the reference (Supplementary Fig. 10)
outliers=~/midProject/data/cohorts/ukbb/hg37/ukb_imp_v3_almarri_hgdp1kgpqc_hg37af_outliers.tsv
plink2=/path/to/plink2
cohort=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc
refname=$(echo $cohort | sed "s/.*\///g");
path=~/midProject/data/cohorts/ukbb/hg37/
hg=37

$plink2 \
--keep-allele-order \
--bfile ${cohort}_hg${hg} --exclude $outliers --make-bed --out ${cohort}_hg${hg}af

$plink \
--keep-allele-order \
--bfile ${cohort}_hg${hg} --exclude $outliers --make-bed --out ${path}ukb_imp_v3_${refname}
```

# Train genetic ancestry RF model

## Perform PCA using plink

Run PCA on the QC'd, outlier-free reference panel, generating variant weights and
allele frequencies needed to later project UKBB samples onto the same PC space.

```{bash pca}
plink2=/path/to/plink2
hg="_hg37"
training=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc${hg}af

### STEP 1: define training and test plinks
pcs=6

### STEP2: PCA
name_training=$(echo "$training" | sed 's:.*/::');
output_training=~/midProject/results/raw_outputs/pca/${name_training};
# Generate var PC weights, scores, and frequency
$plink2 \
--bfile $training \
--keep-allele-order \
--freq \
--pca ${pcs} biallelic-var-wts meanimpute \
--out ${output_training}_${pcs}pcs;
```

### Plot PCA

Plot the reference panel's PCs, coloring by population/study, to visually confirm
the expected population structure.

```{r plot-pca}
library(dplyr)
library(ggplot2)

colors <- c("#F0E442",  # Yellow
            "#CC79A7",  # Pink
            "#E69F00",  # Orange
            "#56B4E9",  # Sky Blue
            "#D55E00",  # Vermillion
            "#0072B2",  # Blue
            "#009E73"  # Green
)

training="almarri_hgdp1kgpqc_hg37af" 
pcs=6
pca <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", training, pcs,"pcs.eigenvec"), header = T, sep = "\t")
colnames(pca)[1] <- "pop"
colnames(pca)[2] <- "s"

plot_pcs <- paste0("PC", seq(pcs))

## add study name
pca <- pca %>%
  mutate(
    study = case_when(
      grepl("^HG", s) | grepl("^LP", s) ~ "HGDP-1KGP",
      grepl("^NA", s) | grepl("^SS", s) ~ "HGDP-1KGP",
      TRUE ~ "Almarri"
    )
  )
  
# Assign pop names to your colors
pop.levels <- unique(pca$pop)
colors_named <- setNames(colors[1:length(pop.levels)], pop.levels)
pca$color_group <- pca$pop

plot_pca_projection <- function(pca, first_pc, second_pc) {
  ggplot(data = pca, aes_string(x = first_pc, y = second_pc, label = "pop", 
                                colour = "color_group", shape = "study")) +
    geom_point(size = 2, stroke = 1, alpha = 0.5) +
    scale_color_manual(
      name = element_blank(),
      values = colors_named
    ) +
    scale_shape_manual(values = c(0, 1, 2)) +
    
    labs(
      x = first_pc,
      y = second_pc
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "right"
    )
}

p<- plot_pca_projection(pca, "PC1", "PC2")
```

### Plot PC loadings and chromosome positions

Plot per-variant PC loadings against chromosome position to check for PCs driven
by localized regions (e.g. inversions or long-range LD) rather than genome-wide
structure.

```{r plot-pc-loadings}
library(ggplot2)
library(gridExtra)
library(dplyr)

training="almarri_hgdp1kgpqc_hg37af"
pcs=6
pcload <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", training, "_", pcs,"pcs.eigenvec.var"), header = T, sep = "\t")
pcload[, paste0("PC", 1:pcs)] <- abs(pcload[, paste0("PC", 1:pcs)])
pcload <- pcload[apply(pcload[, paste0("PC", 1:pcs)] >= 0.1, 1, all), ]
pcload$BP <- as.numeric(sapply(strsplit(as.character(pcload$ID), ":"), `[`, 2))

plots <- list()

# Loop through PCs from 1 to 10
for (pc in 1:pcs) {
  pc_name <- paste0("PC", pc)  # Generate column name dynamically
  
  # Create the plot
  p <- ggplot(pcload, aes(x = BP, y = .data[[pc_name]], color = as.factor(X.CHROM))) +
  geom_point(alpha = 0.6, size = 0.5) +  # Using geom_point instead of jitter for clarity
  scale_color_manual(values = rep(c("black", "grey"), length(unique(pcload$X.CHROM)))) + 
  facet_wrap(~X.CHROM, scales = "free_x", nrow = 1, strip.position = "bottom") +  # Move facet labels to the bottom
  labs(y = paste0(pc_name, " Loadings"), x = "Chromosome") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),   # Removes the background grid
    axis.ticks.x = element_blank(), # Removes x-axis ticks
    axis.text.x = element_blank()   # Removes x-axis labels
  )

  plots[[pc]] <- p
}

# Arrange plots in a grid
grid_plot <- marrangeGrob(plots, nrow = pcs/2, ncol = 2)

# Save the plot
ggsave(filename = paste0("~/midProject/results/figures/", training, "_", pcs,"pcs_manhattan.png"),
       plot = grid_plot, width = 12, height = 16, dpi = 300)
```

## Train RF model

Train a random forest classifier on the reference panel's PCs to predict
population labels, following gnomAD's ancestry-assignment approach.

- gnomAD parameters used as reference points:
  - v2: 6 PCs, 0.9 probability threshold (applied on UKBB)
  - v3: 16 PCs, 0.75 probability threshold (applied on AoU)

```{bash train-rf-wrapper}
train=~/midProject/scripts/analysis/ancestryClassification/02_trainRF.py
pcs=6
prob=0.9
hg="_hg37af"
scores=~/midProject/results/raw_outputs/pca/almarri_hgdp1kgpqc${hg}_${pcs}pcs.eigenvec

python3.9 $train $scores $prob
```

**02_trainRF.py**
```{python train-rf}
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import confusion_matrix, classification_report
from gnomad.sample_qc.ancestry import assign_population_pcs, pc_project
import pandas as pd
import numpy as np
import os
import pickle
import time
import sys


##############################
## Paths
start = time.time()
scores_path = os.path.expanduser(sys.argv[1])
base_name = os.path.splitext(os.path.basename(scores_path))[0]
print(base_name)
model_path = os.path.join(
        os.path.expanduser("~/midProject/results/raw_outputs/ancestryPredictions"),
        f"{base_name}_rf.pkl")
report_path = os.path.join(
        os.path.expanduser("~/midProject/results/raw_outputs/ancestryPredictions"),
        f"{base_name}_rf_report.txt")
        
print(f"2. Loaded training data {(time.time() - start)/60:.2f} minutes.")

##############################
### Load training data
print(time.time(), "READING SCORES")
train_scores = pd.read_csv(scores_path, sep="\t")
pc_cols = [col for col in train_scores.columns if col.startswith("PC")]
train_scores = train_scores.rename(columns={"IID": "s"}) #to prevent the error from gnomad code
train_scores = train_scores.rename(columns={"#FID": "POP"})

prob_thresh=float(sys.argv[2])
##############################
### Train RF model
start = time.time()
df, rf_model = assign_population_pcs(
    train_scores,
    pc_cols=pc_cols,
    min_prob=prob_thresh,
    known_col="POP"
) 

print(f"3. Trained RandomForest in {(time.time() - start)/60:.2f} minutes.")

##############################
### Save the model
print(time.time(), "SAVING THE MODEL")
with open(model_path, "wb") as f:
    pickle.dump(rf_model, f)

##############################
### Save accuracy and model parameters
start = time.time()

# Convert Hail table to pandas for evaluation
#all_labels = list(rf_model.classes_)
try:
    all_labels = list(rf_model.classes_)
    all_labels.append("oth")
except ValueError:
    all_labels = list(rf_model.classes_)  # skip appending "oth"

# Confusion matrix
er_mat = confusion_matrix(df['POP'], df['pop'], labels=all_labels)
er_df = pd.DataFrame(er_mat, index=all_labels, columns=all_labels)

# Classification report
report = classification_report(df['POP'], df['pop'], target_names=all_labels)

# Save in report
with open(report_path, "w") as f:
    f.write("Error rate:\n")
    f.write(er_df.to_string())
    f.write("\n\nRandomForestClassifier Parameters:\n")
    f.write(str(rf_model.get_params()))
    f.write("\n\nClassification Report:\n")
    f.write(report)

print(f"5. Evaluation and outputs saved in {(time.time() - start)/60:.2f} minutes.")
```

# Genetic ancestry prediction

## UKBB ancestry prediction (hg37 plink)

Project UKBB samples onto the reference panel's PC space using `--score`, then run
the trained RF model to predict ancestry.

```{bash apply-ukbb}
plink=/path/to/plink_v1.9/plink
plink2=/path/to/plink2
predict=~/midProject/scripts/analysis/ancestryClassification/03_predict.py
hg="_hg37af"
training=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc${hg}
name_training=$(echo $training | sed "s/.*\///g"); 

pcs=6;
pc_col=$((4 + pcs));

test=~/midProject/data/cohorts/ukbb/hg37/ukb_imp_v3_${name_training}
name_test=ukb_imp_v3
output_test=~/midProject/results/raw_outputs/pca/${name_test}X${name_training}_${pcs}pcs
loadings=~/midProject/results/raw_outputs/pca/${name_training}_${pcs}pcs.eigenvec.var;
freq=~/midProject/results/raw_outputs/pca/${name_training}_${pcs}pcs.afreq;


## (1) score
$plink2 \
--bfile ${test} \
--read-freq $freq \
--score $loadings 2 3 header-read variance-standardize \
--score-col-nums 5-${pc_col} \
--out ${output_test};

## (2) predict ancestry
module load python/3.12.2
module load anaconda3/2024.06
conda activate hail

cd ~/midProject/scripts/analysis/ancestryClassification
python3.9 $predict ${name_training}_${pcs}pcs $name_test

```

**03_predict.py (Plink scores)**
```{python predict}
from sklearn.ensemble import RandomForestClassifier
from sklearn.preprocessing import LabelEncoder
from sklearn.metrics import classification_report, confusion_matrix
import pandas as pd
import numpy as np
import os
import pickle
import time
import sys

##############################
### Paths

training_name = sys.argv[1]
test_name = sys.argv[2]
prob_thresh=0.5

# Build paths based on input names
test_scores_path = os.path.expanduser(
    f"~/midProject/results/raw_outputs/pca/{test_name}X{training_name}.sscore"
)

var_path = os.path.expanduser(
    f"~/midProject/results/raw_outputs/pca/{test_name}X{training_name}.sscore.vars"
)

model_path = os.path.expanduser(
    f"~/midProject/results/raw_outputs/ancestryPredictions/{training_name}.scores_rf.pkl"
)

prob_path = os.path.expanduser(
    f"~/midProject/results/raw_outputs/ancestryPredictions/{test_name}X{training_name}_prob.tsv"
)

report_path = os.path.expanduser(
    f"~/midProject/results/raw_outputs/ancestryPredictions/{test_name}X{training_name}_predictionsReport.txt"
)
##############################
### Read projected scores
print(time.time(), "READING SCORES")
test_scores = pd.read_csv(test_scores_path, sep="\t")
var = pd.read_csv(var_path, sep="\t", header=None)
##############################
# read variance for scaling the scores

n_sscore_vars = len(var)
denoms = np.sqrt(n_sscore_vars)

pc_cols = [col for col in test_scores.columns if col.startswith("PC")]
test_pcs = test_scores[pc_cols].values / denoms
test_labels = test_scores['#FID']
iid = test_scores['IID']

##############################
### Load model
print(time.time(), "LOADING THE MODEL")
# pickle model
with open(model_path, "rb") as f:
    rf = pickle.load(f)

### Predict
print(time.time(), "PREDICTING")

probs = rf.predict_proba(test_pcs)

# Get the class with highest probability
max_probs = np.max(probs, axis=1)
predicted_classes = rf.classes_[np.argmax(probs, axis=1)]

# rf.predict() assigns ancestry based on highest mean probability.
# Apply threshold: if no prob > threshold, set to 'oth'
preds = np.where(max_probs > prob_thresh, predicted_classes, 'oth')

##############################
### Save probabilities
class_labels = rf.classes_
probs_df = pd.DataFrame(probs, columns=class_labels)
probs_df.insert(0, "IID", iid)
probs_df.insert(0, "FID", test_labels)
probs_df["Predicted_pop"] = preds

probs_df.to_csv(prob_path, sep="\t", index=False)

### Save classification report
# Get unique labels from either predicted or true labels
all_labels = sorted(set(test_labels) | set(preds))

# Confusion matrix
er_mat = confusion_matrix(test_labels, preds, labels=all_labels)
er_df = pd.DataFrame(er_mat, index=all_labels, columns=all_labels)

# Classification report
report = classification_report(test_labels, preds, target_names=all_labels)

# Save in report
with open(report_path, "w") as f:
    f.write("Error rate:\n")
    f.write(er_df.to_string())
    f.write("\n\nRandomForestClassifier Parameters:\n")
    f.write(str(rf.get_params()))
    f.write("\n\nClassification Report:\n")
    f.write(report)
```

## MID ancestry refinement

Re-run PCA on predicted-MID UKBB samples alone (unrelated set), project related
MID samples onto that PC space, then use centroid distance to refine which
samples are confidently MID (removing residual outliers not caught by the RF
step).

```{bash mid-refinement}
path=~/midProject/data/cohorts/ukbb/hg37/mid/new_predictions/
plink=/path/to/plink_v1.9/plink
plink2=/path/to/plink2
hg="_hg37af"
training=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc${hg}
name_training=$(echo $training | sed "s/.*\///g"); 
pcs=6;

## STEP1: Re-run PCA for each initial continental category (without reference)
# 1.1 filter in MID samples
# 1.2 run 10 PCs (on all markers or intersecting markers?)
predictions=~/midProject/results/raw_outputs/ancestryPredictions/ukb_imp_v3X${name_training}_${pcs}pcs_prob.tsv
mid_iid=~/midProject/data/cohorts/ukbb/hg37/mid/new_predictions/iid_mid_${name_training}_${pcs}pcs.txt
cohort=~/midProject/data/cohorts/ukbb/hg37/mid/new_predictions/midukb_imp_v3${name_training}_${pcs}pcs
ukbb_king=/path/to/ukbiobank/relatedness/king.kin0

# get MID IIDs
less $predictions | awk -v OFS="\t" '$10 == "MID"' | cut -f 2 | sed "s/_/\t/g" | cut -f 1 > $mid_iid

# keep predicted MID samples and join chr
for i in {1..22}; do
file=/path/to/ukbiobank/imputation/bgen_to_pgen/ukb_imp_chr${i}_v3;
output=~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_chr${i}_v3;
$plink2 \
--pfile ${file} --snps-only 'just-acgt' \
--set-all-var-ids chr@:#:\$r:\$a \
--keep-fam $mid_iid --make-bed --out $output; done

# merge chr
> ${path}mergelist.txt
for i in {1..22}; do
echo ~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_chr${i}_v3 >> ${path}mergelist.txt;
done

$plink \
--merge-list ${path}mergelist.txt --make-bed --out $cohort
#rm midukb_imp_chr*_v3*

# get relateds
grep -Ff $mid_iid ${ukbb_king} | cut -f 3 > ${path}mid_relateds.txt

# related and unrelated
$plink \
--bfile ${cohort} --remove-fam mid_relateds.txt \
--make-bed --out ${cohort}_unrel

$plink \
--bfile ${cohort} --keep-fam mid_relateds.txt \
--make-bed --out ${cohort}_rel

# QC
$plink \
--bfile ${cohort}_unrel --geno 0.1 --make-bed --out ${cohort}_unrelqc

$plink \
--bfile ${cohort}_unrelqc --mind 0.1 \
--make-bed --out ${cohort}_unrelqc

$plink \
--bfile ${cohort}_unrelqc --maf 0.01 --make-bed --out ${cohort}_unrelqc

#remove MHC
awk -F'\t' '$1 == 6 {split($2, pos, ":"); if (pos[2] >= 28510120 && pos[2] <= 33480577) print}' ${cohort}_unrelqc.bim | cut -f 2 > ${cohort}_unrelqc.mhc
$plink2 \
--bfile ${cohort}_unrelqc --exclude ${cohort}_unrelqc.mhc --keep-allele-order --make-bed --out ${cohort}_unrelqc

# LD pruning
$plink2 \
--bfile ${cohort}_unrelqc --indep-pairwise 50 5 0.5 --keep-allele-order --out ${cohort}_unrelqc

# Apply pruning 
$plink2 --bfile ${cohort}_unrelqc \
  --extract ${cohort}_unrelqc.prune.in \
  --keep-allele-order \
  --make-bed \
  --out ${cohort}_unrelqc
  
# PCA
cd ~/midProject/results/raw_outputs/pca
pcs=10
# Generate var PC weights, scores, and frequency
$plink2 \
--bfile ${cohort}_unrelqc \
--keep-allele-order \
--freq \
--pca ${pcs} biallelic-var-wts \
--out ${cohort}_unrelqc${pcs}pcs

#project relateds
pc_col=$((4 + pcs))
# Use --score to compute the necessary dot products with the variant weights
$plink2 \
--bfile ${cohort}_rel \
--read-freq ${cohort}_unrelqc${pcs}pcs.afreq \
--score ${cohort}_unrelqc${pcs}pcs.eigenvec.var 2 3 header-read no-mean-imputation variance-standardize \
--score-col-nums 5-${pc_col} \
--out ${cohort}_rel${pcs}pcs
```

## Centroid distance calculation

Compute each sample's distance from the population centroid across the first N
PCs, and use that distance to iteratively choose a cutoff that excludes outliers
from the "MID-like" set (functions adapted from the
[UKBB pan-ancestry project](https://github.com/atgu/ukbb_pan_ancestry)).

```{r centroid-distance}
## STEP2: centroid calculation
library(RColorBrewer)
library(cowplot)
library(dplyr)
library(ggplot2)

unrel <- read.csv("~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_v3almarri_hgdp1kgp_trainingqc_pruned6pc_unrelqc_pruned10pc.eigenvec", header = T, sep = "\t")
var <- read.csv("~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_v3almarri_hgdp1kgp_trainingqc_pruned6pc_unrelqc_pruned10pc.eigenval", header = F)
pcs=6
scores <- read.csv("~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_v3almarri_hgdp1kgp_trainingqc_pruned6pc_rel10pc.sscore", header = T, sep = "\t")
rel <- data.frame(X.FID = scores$X.FID, IID = scores$IID)

for (i in 1:10) {
  pc_col <- paste0("PC", i)
  
  # Get the variance value from the row name
  var_val <- var[i, 1]  # assuming the variance is in the first (or only) column
  denom <- -sqrt(var_val) / 2
  # Apply transformation
  rel[[pc_col]] <- scores[[paste0(pc_col, "_AVG")]] / denom
}

mid <- rbind(rel, unrel) 
rm(unrel, rel, scores, denom, i, pc_col, var_val) 

# functions adapted from https://github.com/atgu/ukbb_pan_ancestry/blob/0835f603879aed24b4ef7fa3d44f63e16f34194e/plot_ukbb_continental_pca.R#L423
# function to generate distance of each individual to the centroid
pop_ellipse <- function(df, num_ellipses) {
  # get mean and SD of each PC among each pop
  pc_nams <- paste("PC",1:10,sep="")
  mean_pcs <- colMeans(df[,pc_nams])
  sd_pcs <- apply(df[,pc_nams],2,sd)
  # compute centroid distance for each individual
  centroid_dist <- rep(0,nrow(df))
  for(i in 1:num_ellipses) {
    centroid_dist <- centroid_dist + (df[,pc_nams[i]]-mean_pcs[i])^2/(sd_pcs[i]^2)
  }
  pop_dist <- df %>%
    mutate(centroid_dist=centroid_dist)
  return(pop_dist)
}

# function to exclude outliers
pop_centroid <- function(ind_dist, cutpoint0, cutpoint1) {
  pop_cut <- subset(ind_dist, centroid_dist < cutpoint0)
  p_centroid <- ggplot(pop_cut, aes(x=centroid_dist)) + 
    geom_histogram(bins=50) + 
    labs(title=paste0('Sample size: ', nrow(subset(ind_dist, centroid_dist < cutpoint0)), ' -> ', nrow(subset(ind_dist, centroid_dist < cutpoint1)))) +
    xlab('Centroid distance') + 
    ylab('Count') +
    geom_vline(xintercept=cutpoint1) +
    theme_bw() +
    theme(text = element_text(size=16))
  return(list(p=p_centroid, pop_cut=pop_cut))
}
plot_pca_density <- function(dataset, first_pc, second_pc) {
  pc_biplot <- ggplot(dataset, aes_string(x=first_pc, y=second_pc)) +
    geom_hex(bins=50) +
    scale_fill_gradientn(trans = "log", breaks=c(1,20,400,8000,163000), name='Count',
                         colours = rev(brewer.pal(5,'Spectral'))) +
    theme_classic() +
    theme(text = element_text(size=16))
  return(pc_biplot)
}
save_filt_plots <- function(pop_name, pop_dist, cutpoint0, cutpoint1) {
  p_centroid0 = pop_centroid(pop_dist, cutpoint0, cutpoint1)
  ggsave(paste0(pop_name, '_within_pop_centroid_nofilt.pdf'), p_centroid0$p, height=7, width=7)
  p2 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint0), 'PC1', 'PC2')
  ggsave(paste0(pop_name, '_within_pop_nofilt_pc1_2.png'), p2)
  p3 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint0), 'PC3', 'PC4')
  ggsave(paste0(pop_name, '_within_pop_nofilt_pc3_4.png'), p3)
  p4 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint0), 'PC5', 'PC6')
  ggsave(paste0(pop_name, '_within_pop_nofilt_pc5_6.png'), p4)
  p_centroid1 = pop_centroid(pop_dist, cutpoint1, cutpoint1)
  ggsave(paste0(pop_name, '_within_pop_centroid_filt.pdf'), p_centroid1$p, height=7, width=7)
  p6 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint1), 'PC1', 'PC2')
  ggsave(paste0(pop_name, '_within_pop_filt_pc1_2.png'), p6)
  p7 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint1), 'PC3', 'PC4')
  ggsave(paste0(pop_name, '_within_pop_filt_pc3_4.png'), p7)
  p8 <- plot_pca_density(subset(pop_dist, centroid_dist < cutpoint1), 'PC5', 'PC6')
  ggsave(paste0(pop_name, '_within_pop_filt_pc5_6.png'), p8)
  my_plot=plot_grid(p_centroid0$p, p2, p3, p4, p_centroid1$p, p6, p7, p8, nrow=2)
  save_plot(paste0(pop_name, '_within_pop.png'), my_plot, base_height=10, base_width = 18)
  return(p_centroid1$pop_cut)
}

# UKBB cutoff
mid_cut <- save_filt_plots('~/midProject/results/figures/ukbb_cutpoint/mid', mid_dist <- pop_ellipse(mid, 5), 1000, 15)

## save refined MID assignments

write.table(mid_cut, paste0("~/midProject/results/raw_outputs/ancestryPredictions/refinedMID15cutoff_ukb_imp_v3", pcs, "pc.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)

##########################
#Choose new cutoff
d <- pop_ellipse(mid, 5)
hist(d[d$centroid_dist < 100, ]$centroid_dist)

pop_centroid(d, 60, 7.25)
mid_cut <- save_filt_plots('~/midProject/results/figures/new_cutpoint/mid', mid_dist <- pop_ellipse(mid, 5), 1000, 9)
write.table(mid_cut, paste0("~/midProject/results/raw_outputs/ancestryPredictions/refinedMID9cutoff_ukb_imp_v3", pcs, "pc.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)

mid$cut <- ifelse(mid$IID %in% mid_cut$IID, "included", "outlier")

####### plot outliers ##########
#plot pca
plot_pca <- function(dataset, first_pc, second_pc) {
  ggplot(data = dataset, aes_string(x = first_pc, y = second_pc, color = "cut")) +
    geom_point(size = 1, stroke = 1, alpha = 0.2) + 
    theme_classic() +
    theme(
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "right",
      legend.title = element_blank()
    )
}

save_pca_plots <- function(pop_name, pop_dist, cutpoint0, cutpoint1, cutpoint2) {
  p_centroid0 = pop_centroid(pop_dist, 1000, cutpoint0)
  pop_cut = pop_centroid(pop_dist, cutpoint0, cutpoint0)$pop_cut
  pop_dist$cut <- ifelse(mid$IID %in% pop_cut$IID, "included", "outlier")
  p2 <- plot_pca(pop_dist, 'PC1', 'PC2')
  p3 <- plot_pca(pop_dist, 'PC3', 'PC4')
  p4 <- plot_pca(pop_dist, 'PC5', 'PC6')
  p_centroid1 = pop_centroid(pop_dist, 1000, cutpoint1)
  pop_cut = pop_centroid(pop_dist, cutpoint1, cutpoint1)$pop_cut
  pop_dist$cut <- ifelse(mid$IID %in% pop_cut$IID, "included", "outlier")
  p6 <- plot_pca(pop_dist, 'PC1', 'PC2')
  p7 <- plot_pca(pop_dist, 'PC3', 'PC4')
  p8 <- plot_pca(pop_dist, 'PC5', 'PC6')
  p_centroid2 = pop_centroid(pop_dist, 1000, cutpoint2)
  pop_cut = pop_centroid(pop_dist, cutpoint2, cutpoint2)$pop_cut
  pop_dist$cut <- ifelse(mid$IID %in% pop_cut$IID, "included", "outlier")
  p10 <- plot_pca(pop_dist, 'PC1', 'PC2')
  p11 <- plot_pca(pop_dist, 'PC3', 'PC4')
  p12 <- plot_pca(pop_dist, 'PC5', 'PC6')
  my_plot=plot_grid(p_centroid0$p, p2, p3, p4, p_centroid1$p, p6, p7, p8, p_centroid2$p, p10, p11, p12, nrow=3)
  save_plot(paste0(pop_name, '_cutpoints.png'), my_plot, base_height=15, base_width = 20)
}

save_pca_plots('~/midProject/results/figures/new_cutpoint/mid', mid_dist <- pop_ellipse(mid, 5), 30, 9, 3)
```

## Plot PCA projection on training data

Compare the MID samples predicted in this study vs. those predicted directly in
UKBB, projected onto the reference panel's PCA space.

```{r plot-pca-projection}
library(dplyr)
library(ggplot2)
library(colorspace)
library(patchwork)
library(stringr)
# plot AF

training="almarri_hgdp1kgpqc_hg37af"
test="ukb_imp_v3"
pcs=6
cutoff=7

pca <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", training, "_", pcs,"pcs.eigenvec"), header = T, sep = "\t")
projection <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", test, "X", training, "_", pcs,"pcs.sscore"), header = T, sep = "\t")
var <- read.csv(paste0("~/midProject/results/raw_outputs/pca/", training, "_", pcs, "pcs.eigenval"), header = F, sep = "\t")
#read refined predictions
prediction <- read.csv(paste0("~/midProject/results/raw_outputs/ancestryPredictions/", test, "X", training, "_", pcs,"pcs_prob.tsv"), header = T, sep = "\t")
refined <- read.csv(paste0("~/midProject/results/raw_outputs/ancestryPredictions/refinedMID7cutoff_ukb_imp_v3", pcs, "pcs.tsv"), header = T, sep = "\t")

###

colors <- c("#F0E442",  # Yellow
            "#CC79A7",  # Pink
            "#E69F00",  # Orange
            "#56B4E9",  # Sky Blue
            "#0072B2",  # Blue
            "#009E73",  # Green
            "#D55E00",  # Vermillion
            "#999999" # grey
)
light_colors <- lighten(colors, amount = 0.8)

## scale projected scores  
for (i in 1:pcs) {
  pc_col <- paste0("PC", i, "_AVG")
  # Get the variance value from the row name
  var_val <- var[i, 1]
  denom <- -sqrt(var_val)/2
  # Apply transformation
  projection[[paste0("PC", i)]] <- projection[[pc_col]] / denom
}

prediction$IID <- sapply(strsplit(prediction$IID, "_"), `[`, 1)
projection$IID <- sapply(strsplit(projection$IID, "_"), `[`, 1)

#Keep only MID in predictions
prediction <- prediction[(prediction$FID == "MID" | prediction$Predicted_pop == "MID"), ]
count(prediction[prediction$Predicted_pop=="MID", ])
prediction$Predicted_pop <- ifelse(prediction$IID %in% refined$IID, "MID", "Outlier")
count(prediction[prediction$Predicted_pop=="MID", ])

# join projection scores to MID predicted samples
count(projection)
projection <- projection[projection$IID %in% prediction$IID, ]
count(projection)
projection$prediction <- prediction$Predicted_pop[match(projection$IID, prediction$IID)]

## add study name
pca <- pca %>%
  mutate(
    study = case_when(
      grepl("^HG", IID) | grepl("^LP", IID) ~ "HGDP-1KGP",
      grepl("^NA", IID) | grepl("^SS", IID) ~ "HGDP-1KGP",
      TRUE ~ "Almarri"
    )
  )
projection$study <- "UKBB"

# empty df for legend
mid_legend <- data.frame(
  mid_label = factor(c("UKBB only", "This study only", "Both"),
                     levels = c("UKBB only", "This study only", "Both"))
)

## plot
plot_pca <- function(pca, first_pc, second_pc) {
  ggplot(pca, aes_string(x = first_pc, y = second_pc, colour = "X.FID", shape = "study")) +
    geom_point(size = 2, stroke = 1, alpha = 0.8) +
    scale_color_manual(values = light_colors) +
    scale_shape_manual(values = c(0, 1, 4)) +
    
    # UKBB only
    geom_point(
      data = projection[projection$X.FID == "MID" & projection$prediction != "MID", ],
      aes_string(x = first_pc, y = second_pc, shape = "study"),
      inherit.aes = FALSE,
      color = "#888888", size = 1, alpha = 1
    ) +
    
    # This study only
    geom_point(
      data = projection[projection$X.FID != "MID" & projection$prediction == "MID", ],
      aes_string(x = first_pc, y = second_pc, shape = "study"),
      inherit.aes = FALSE,
      color = "#aa4499", size = 1, alpha = 1
    ) +
    
    # Both
    geom_point(
      data = projection[projection$X.FID == "MID" & projection$prediction == "MID", ],
      aes_string(x = first_pc, y = second_pc, shape = "study"),
      inherit.aes = FALSE,
      color = "#661103", size = 1, alpha = 0.7
    ) +
    
    # Legend-only layer
    geom_point(
      data = mid_legend,
      aes(x = 0, y = 0, fill = mid_label),
      shape = 16, color = "white", size = 3,
      alpha = 0, show.legend = TRUE,
      inherit.aes = FALSE
    ) +
    
    scale_fill_manual(
      name = "MID prediction",
      values = c(
        "UKBB only" = "#888888",
        "This study only" = "#aa4499",
        "Both" = "#661103"
      )
    ) +
    
    guides(
      fill = guide_legend(override.aes = list(shape = 21, alpha = 1)),
      color = guide_legend(override.aes = list(shape = 16)),
      shape = guide_legend()
    ) +
    labs(
      x = paste0(first_pc, " (", round(var[str_extract(first_pc, "\\d+"), 1]/sum(var$V1)*100, 2),"%)"),
      y = paste0(second_pc, " (", round(var[str_extract(second_pc, "\\d+"), 1]/sum(var$V1)*100, 2),"%)")
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "right",
      legend.title = element_blank()
    )
}

## plot multiple PCs and save
p1 <- plot_pca(pca, "PC1", "PC2")
ggsave(plot=p1, paste0("~/midProject/results/figures/ukbb_cutpoint/pc1pc2mid_", training, pcs, "pc.pdf"), width = 8, height = 6)
p2 <- plot_pca(pca, "PC3", "PC4")
ggsave(plot=p2, paste0("~/midProject/results/figures/ukbb_cutpoint/pc3pc4mid_", training, pcs, "pc.pdf"), width = 8, height = 6)
p3 <- plot_pca(pca, "PC5", "PC6")
ggsave(plot=p3, paste0("~/midProject/results/figures/ukbb_cutpoint/pc5pc6mid_", training, pcs, "pc.pdf"), width = 8, height = 6)

p1 <- p1 + theme(legend.position = "none")
p2 <- p2 + theme(legend.position = "none")
p3 <- p3 + theme(legend.position = "right")
# combine them horizontally
combined <- p1 + p2 + p3 + plot_layout(guides = "collect")
ggsave(plot=combined, paste0("~/midProject/results/figures/ukbb_cutpoint/pc1to6mid_", training, pcs, "pc.png"), width = 15, height = 5)

plot_pca(pca, "PC2", "PC6")
ggsave(paste0("~/midProject/results/figures/ukbb_cutpoint/pc2pc6mid_", training, pcs, "pc.pdf"), width = 8, height = 6)
```

# MID-like set validation

## Compare place of birth

```{bash compare-birthplace}
cd ~/midProject/data/cohorts/ukbb/hg37/birth_place 
newpred=~/midProject/results/raw_outputs/ancestryPredictions/refinedMID7cutoff_ukb_imp_v36pcs.tsv
ukbbpred=~/midProject/results/raw_outputs/ancestryPredictions/ukb_imp_v3Xalmarri_hgdp1kgpqc_hg37af_6pcs_prob.tsv

# add label to country key
awk 'NR==FNR {map[$1]=$2; next} {print $0 "\t" map[$2]}' countryofbirth_nonuk_labels.tsv countryofbirth_nonUK.txt > countryofbirth_nonuk_key_labels.txt

##get ID of both sets
less $newpred | cut -f 1 | grep -v FID > newmid.txt
less $ukbbpred | grep ^MID | cut -f 2 | grep -v IID | sed "s/_/\t/g" | cut -f 1 > ukbbmid.txt

# add country of birth code
sample=new
sample=ukbb
grep -Ff ${sample}mid.txt countryofbirth_nonuk_key_labels.txt | sed "s/ /\t/g" > ${sample}mid_country_nonUK.txt
```

### Plot place of birth comparison

Map predicted MID samples by self-reported country of birth, comparing UKBB-only,
this-study-only, and overlapping predictions.

```{r plot-birthplace}
# Load libraries
library(data.table)
library(dplyr)
library(ggplot2)
library(maps)
library(countrycode)
library(scatterpie)
library(patchwork)


# --- Input file paths ---
output = "~/midProject/results/figures/rf_model/ukbb_birth/"
file0 <- "~/midProject/data/cohorts/ukbb/hg37/birth_place/countryofbirth_nonuk_key_labels.txt" 
file1 <- "~/midProject/data/cohorts/ukbb/hg37/birth_place/ukbbmid_country_nonUK.txt"
file2 <- "~/midProject/data/cohorts/ukbb/hg37/birth_place/newmid_country_nonUK.txt"

# --- Read data ---
all <- read.table(file0, sep="\t", header=TRUE, stringsAsFactors=FALSE, col.names=c("id", "code", "country"))
ukbb <- read.table(file1, sep="\t", header=FALSE, stringsAsFactors=FALSE, col.names=c("id", "code", "country"))
new  <- read.table(file2, sep="\t", header=FALSE, stringsAsFactors=FALSE, col.names=c("id", "code", "country"))

# --- Assign source labels ---
all <- na.omit(all)
ukbb <- na.omit(ukbb)
new <- na.omit(new)

# --- combine datasets ---
all$src  <- "all"
ukbb$src <- "ukbb"
new$src  <- "new"

all_dt  <- as.data.table(all)
ukbb_dt <- as.data.table(ukbb)
new_dt  <- as.data.table(new)

# --- identify overlaps ---
both_ids <- intersect(ukbb_dt$id, new_dt$id)

# --- counts per country ---
counts_all  <- all_dt[, .N, by = country]
counts_ukbb <- ukbb_dt[, .N, by = country]
counts_new  <- new_dt[, .N, by = country]

counts_both <- all_dt[id %in% both_ids, .N, by = country]

# merge
counts <- Reduce(function(x, y) merge(x, y, by = "country", all = TRUE),
                 list(counts_all, counts_ukbb, counts_new, counts_both))

colnames(counts)[2:5] <- c("all","ukbb","new","both")
counts[is.na(counts)] <- 0

# subtract overlaps so pies sum correctly
counts[, ukbb_only := ukbb - both]
counts[, new_only  := new  - both]
counts[, none      := all - ukbb_only - new_only - both]

# --- map data ---
world <- map_data("world")
country_centroids <- world %>%
  group_by(region) %>%
  summarise(long = median(range(long)), lat = median(range(lat)))

# --- merge coordinates ---
plot_df <- merge(counts, country_centroids,
                 by.x = "region", by.y = "region", all.x = TRUE)

# round coordinates
precision <- 3

plot_df <- na.omit(plot_df) %>% 
  mutate(
    long_round = round(long / precision) * precision,
    lat_round = round(lat / precision) * precision
  ) %>% 
  group_by(long_round, lat_round) %>% 
  summarise(
    long = first(long),
    lat = first(lat),
    all = sum(all, na.rm = TRUE),
    ukbb_only = sum(ukbb_only, na.rm = TRUE),
    new_only = sum(new_only, na.rm = TRUE),
    both = sum(both, na.rm = TRUE),
    none = sum(none, na.rm = TRUE),
    countries = paste(unique(country), collapse = ", "),
    region = first(region),
    .groups = "drop"
  )

# --- add continent ---
lon_range <- c(-20, 55)
lat_range <- c(8, 44)

plot_df$continent_group <- dplyr::case_when(
  plot_df$long_round >= lon_range[1] & plot_df$long_round <= lon_range[2] &
    plot_df$lat_round >= lat_range[1] & plot_df$lat_round <= lat_range[2] ~ "MID",
  plot_df$long_round < -20 ~ "Americas",
  plot_df$lat_round < 8 & plot_df$long_round < 55 ~ "Africa",
  plot_df$lat_round > 44 & plot_df$long_round < 60 & plot_df$long_round > -24 ~ "Europe",
  TRUE ~ "Asia"
)

# --- plotting function ---
make_map <- function(d) {
  if (nrow(d) == 0) return(ggplot() + theme_void())
  
  xlim <- range(d$long, na.rm = TRUE)
  ylim <- range(d$lat,  na.rm = TRUE)
  
  ggplot() +
    geom_polygon(data = world,
                 aes(x = long, y = lat, group = group),
                 fill = "grey95", color = NA) +
    
    geom_scatterpie(
      data = d,
      aes(x = long_round, y = lat_round, r = sqrt(all)/5),
      cols = c("none","ukbb_only","new_only","both"),
      color = "black",
      lwd = 0.3 
    ) +
    
    scale_fill_manual(values = c(
      "none" = "white",
      "ukbb_only" = "#888888",
      "new_only" = "#aa4499",
      "both" = "#661103"
    ),
    labels = c("UKBB participants","MID inferred by UKBB","MID inferred by this study","MID inferred by both"),
    name = "Dataset") +
    
    coord_fixed(
      xlim = xlim + c(-2, 2),
      ylim = ylim + c(-2, 2),
      clip = "on"
    ) +
    
    theme_void() +
    theme(plot.margin = margin(0, 0, 0, 0))
}

# --- create maps ---
mid      <- make_map(dplyr::filter(plot_df, continent_group == "MID"))
map_afr  <- make_map(dplyr::filter(plot_df, continent_group == "Africa"))
map_amr  <- make_map(dplyr::filter(plot_df, continent_group == "Americas"))
map_asia <- make_map(dplyr::filter(plot_df, continent_group == "Asia"))
map_eur  <- make_map(dplyr::filter(plot_df, continent_group == "Europe"))

# --- combine ---
combined_plot <-
  (mid) /
  (map_afr + map_amr) /
  (map_asia + map_eur) +
  plot_layout(
    guides = "collect",
    heights = c(3, 2.5, 1.5)
  ) &
  theme(
    legend.position = "top",
    plot.margin = margin(2, 2, 2, 2),
    panel.spacing = unit(0.2, "lines")
  )

# --- save ---
ggsave(
  filename = paste0(output, "site_of_birth.pdf"),
  plot = combined_plot,
  width = 15,
  height = 10,
  device = cairo_pdf,
  bg = "white"
)
```

## Compare GWAS

### Extract all MID samples (UKBB predictions and new predictions)

Pool predicted-MID samples across chromosomes, split related/unrelated, run QC
(call rate, MAF, MHC exclusion, LD pruning) to prepare for downstream GWAS.

```{bash extract-mid-samples}
plink=/path/to/plink_v1.9/plink
plink2=/path/to/plink2
hg="_hg37af"
training=~/midProject/data/reference/cohorts/training/almarri_hgdp1kgpqc${hg}
name_training=$(echo $training | sed "s/.*\///g"); 
pcs=6;

## STEP1: Re-run PCA for each initial continental category (without reference)
# 1.1 filter in MID samples
# 1.2 run 10 PCs (on all markers or intersecting markers?)
predictions=~/midProject/results/raw_outputs/ancestryPredictions/ukb_imp_v3X${name_training}_${pcs}pcs_prob.tsv
mid_iid=~/midProject/data/cohorts/ukbb/hg37/mid/all_predictions/iid_mid.txt
cohort=~/midProject/data/cohorts/ukbb/hg37/mid/all_predictions/mid
output_path=~/midProject/data/cohorts/ukbb/hg37/mid/all_predictions/
# get MID IIDs
less $predictions | grep MID | cut -f 2 | sed "s/_/\t/g" | cut -f 1 > $mid_iid

# keep predicted MID samples and join chr
for i in {1..22}; do
file=/path/to/ukbiobank/imputation/bgen_to_pgen/ukb_imp_chr${i}_v3;
output=${output_path}midukb_imp_chr${i}_v3;
$plink2 \
--pfile ${file} --snps-only 'just-acgt' \
--set-all-var-ids chr@:#:\$r:\$a \
--keep-fam $mid_iid --make-bed --out $output; done

# merge chr
> mergelist.txt
for i in {1..22}; do
echo ${output_path}midukb_imp_chr${i}_v3 >> mergelist.txt;
done

$plink \
--merge-list mergelist.txt --make-bed --out $cohort
#rm midukb_imp_chr*_v3*

# get relateds
grep -Ff $mid_iid /path/to/ukbiobank/relatedness/king.kin0 | cut -f 3 > mid_relateds.txt

# related and unrelated
$plink \
--bfile ${cohort} --remove-fam mid_relateds.txt \
--make-bed --out ${cohort}_unrel

$plink \
--bfile ${cohort} --keep-fam mid_relateds.txt \
--make-bed --out ${cohort}_rel

# QC
$plink \
--bfile ${cohort}_unrel --geno 0.1 --make-bed --out ${cohort}_unrelqc

$plink \
--bfile ${cohort}_unrelqc --mind 0.1 \
--make-bed --out ${cohort}_unrelqc

$plink \
--bfile ${cohort}_unrelqc --maf 0.01 --make-bed --out ${cohort}_unrelqc

#remove MHC
awk -F'\t' '$1 == 6 {split($2, pos, ":"); if (pos[2] >= 28510120 && pos[2] <= 33480577) print}' ${cohort}_unrelqc.bim | cut -f 2 > ${cohort}_unrelqc.mhc
$plink2 \
--bfile ${cohort}_unrelqc --exclude ${cohort}_unrelqc.mhc --keep-allele-order --make-bed --out ${cohort}_unrelqc

# LD pruning
$plink2 \
--bfile ${cohort}_unrelqc --indep-pairwise 50 5 0.2 --keep-allele-order --out ${cohort}_unrelqc

# Apply pruning 
$plink2 --bfile ${cohort}_unrelqc \
  --extract ${cohort}_unrelqc.prune.in \
  --keep-allele-order \
  --make-bed \
  --out ${cohort}_unrelqc
```

### Prepare sample files

Extract UKBB-only vs. this-study-only MID sample sets from the pooled MID cohort,
filter by MAF, LD-prune, and compute PCs to use as GWAS covariates.

```{bash prepare-sample-files}
cd ~/midProject/scripts/analysis/gwas 

newpred=~/midProject/results/raw_outputs/ancestryPredictions/refinedMID7cutoff_ukb_imp_v36pcs.tsv
ukbbpred=~/midProject/results/raw_outputs/ancestryPredictions/ukb_imp_v3Xalmarri_hgdp1kgpqc_hg37af_6pcs_prob.tsv
plink_file=~/midProject/data/cohorts/ukbb/hg37/mid/all_predictions/mid_unrelqc

# done once
#cut -f1,2 $ukbbpred | grep ^MID | sed "s/_/\t/g" | cut -f 2 > ukbb_midsamples.txt
#cut -f1 $newpred | grep -v D > new_midsamples.txt

# extract samples, filter by MAF and prune
sample=ukbb; 
sample=new;
$plink2 \
--bfile $plink_file \
--keep-fam ${sample}_midsamples.txt \
--keep-allele-order \
--maf 0.01 --max-maf 0.99 \
--indep-pairwise 50 5 0.2 \
--make-bed --out $sample

# apply pruning
$plink2 \
--bfile $sample \
--keep-allele-order \
--extract ${sample}.prune.in \
--make-bed \
--out $sample

### run pca for covariates
$plink2 \
--bfile $sample \
--freq counts \
--pca approx 4 \
--out $sample
```

### Prepare phenotype file

Phenotype/covariate field definitions are listed in `phenotypes.txt` (columns:
row number, UKBB field code, phenotype name, column index in the raw UKBB table).
Phenotype groups and their associated covariates:

| phenotypes | pheno numbers | covariates | cov numbers |
|---|---|---|---|
| 01 blood | ph:1,11-15,18-20,22-23,25,28-30,32-33 | sex, age, dilution factor, PCs | ph:2,10,34; PCs |
| 02 lipids | ph:16,17,21,24,26,27,31 | sex, age, dilution factor, chol medication, insulin, PC | ph:2,10,34; med:2,4; PCs |
| 03 measurements (Height, BMI) | ph:6,9 | sex, age, PC | ph:2,10; PCs |
| 04 bp | ph:3,4 | sex, age, blood pressure medication, PC | ph:2,10; med:3; PCs |
| 05 self-reported binary | can: 1,2,3,4; ill: 2,3,4 | sex, age, PC | ph:2,10; PCs |

Build the combined phenotype/covariate table, derive medication and
self-reported-condition binary phenotypes, then split into per-group phenotype
and covariate files.

```{bash prepare-phenotypes}
cd ~/midProject/scripts/analysis/gwas 

newpred=~/midProject/results/raw_outputs/ancestryPredictions/refinedMID7cutoff_ukb_imp_v36pcs.tsv
ukbbpred=~/midProject/results/raw_outputs/ancestryPredictions/ukb_imp_v3Xalmarri_hgdp1kgpqc_hg37af_6pcs_prob.tsv
plink_file=~/midProject/data/cohorts/ukbb/hg37/mid/midukb_imp_v3almarri_hgdp1kgpqc_hg37af_6pcs_unrelqc
pheno=/path/to/ukbiobank/main_dataset/ukb678697.tab
map=/path/to/ukbiobank/main_dataset/ukb678697_header.tab
phdir=~/midProject/data/cohorts/ukbb/phenotypes/

##get all phenotypes (see phenotypes.txt for field definitions)
cut -f 2 "$phdir/phenotypes.txt" | paste -sd '\t' - > "${phdir}/all_samples_pheno.txt"
cut -f $(cut -f3 "$phdir/phenotypes.txt" | paste -sd, -) "$pheno" | grep -v eid >> "${phdir}/all_samples_pheno.txt"

# create medications phenotypes
awk -F'\t' '{
  if (NR==1) {
    print $1 "\tCholesterol_lowering_medication\tBlood_pressure_medication\tInsulin";
  } else {
    clm=($5==1?"1":($5=="NA"||$5==-1||$5==-3?"NA":"0"));
    bpm=($5==2?"1":($5=="NA"||$5==-1||$5==-3?"NA":"0"));
    ins=($5==3?"1":($5=="NA"||$5==-1||$5==-3?"NA":"0"));
    print $1 "\t" clm "\t" bpm "\t" ins;
  }
}' "${phdir}/all_samples_pheno.txt" > "${phdir}/medication.txt"

# create self-reported illness phenotypes
awk -F'\t' '{
  if (NR==1) {
    print $1 "\tAsthma\tT2D\tCardiac_disease";
  } else {
    a=($7==1066?"1":($7=="NA"?"NA":"0"));
    b=($7==1111?"1":($7=="NA"?"NA":"0"));
    c=($7==1223?"1":($7=="NA"?"NA":"0"));
    print $1 "\t" a "\t" b "\t" c;
  }
}' "${phdir}/all_samples_pheno.txt" > "${phdir}/illness.txt"

# create self-reported cancer phenotypes
awk -F'\t' '{
  if (NR==1) {
    print $1 "\tBreast\tColorectal\tProstate";
  } else {
    a=($6==1002?"1":($6=="NA"?"NA":"0"));
    b=($6==1020?"1":($6=="NA"?"NA":"0"));
    c=($6==1044?"1":($6=="NA"?"NA":"0"));
    print $1 "\t" a "\t" b "\t" c;
  }
}' "${phdir}/all_samples_pheno.txt" > "${phdir}/cancer.txt"

### STEP 1: create phenotype files
# normalize quantitative phenotypes using R
# create binary phenotypes file
paste "${phdir}/cancer.txt" "${phdir}/illness.txt" | awk -v OFS="\t" '{print $1,$1,$2,$3,$4,$6,$7,$8}' | sed "s/ID/IID/g" | sed "s/^IID/FID/g" > ${phdir}/05_pheno.tsv

### STEP 2: create covariates files
#01 blood
less ${phdir}/all_samples_pheno.txt | cut -f 1,2,10,34 > ${phdir}/01_cov.tsv
# 02 lipids
less ${phdir}/medication.txt | cut -f 2,4 > tmp
paste ${phdir}/01_cov.tsv tmp > ${phdir}/02_cov.tsv
# 03 measurements
less ${phdir}/all_samples_pheno.txt | cut -f 1,2,10 > ${phdir}/03_cov.tsv
# 04 bp
less ${phdir}/medication.txt | cut -f 3 > tmp
paste ${phdir}/03_cov.tsv tmp > ${phdir}/04_cov.tsv
# 05 binary
less ${phdir}/all_samples_pheno.txt | cut -f 1,2,10 > ${phdir}/05_cov.tsv
```

### Run GWAS

For each phenotype group, extract the relevant samples and covariates (adding
PCs), then run `plink2 --glm` for both the UKBB-predicted and this-study-predicted
MID sample sets.

```{bash run-gwas}
cd ~/midProject/scripts/analysis/gwas 
plink2=/path/to/plink2
phdir=~/midProject/data/cohorts/ukbb/phenotypes/
sample=ukbb
sample=new
output=~/midProject/results/raw_outputs/gwas

for i in {1..5}; do
pheno=${phdir}/0${i}_pheno.tsv
cov=${phdir}/0${i}_cov.tsv
n=$(( $(awk '{print NF; exit}' "$cov") + 6 ))

### Extract sample from phenotype file
cut -f 1 ${sample}.fam > tmp
head -n 1 $pheno > ${sample}0${i}_pheno.tsv
grep -f <(sed 's/^/^/' tmp) $pheno >> ${sample}0${i}_pheno.tsv

### Extract sample from covariates and add PCs
head -n 1 $cov > tmp1
grep -f <(sed 's/^/^/' tmp) "$cov" | sort -k1,1n >> tmp1
paste ${sample}.eigenvec tmp1 | cut -f 1-6,8-$n > ${sample}0${i}_cov.tsv
done

# run GWAS for blood biomarkers
for sample in ukbb new; do
for i in {1..5}; do
pheno=${sample}0${i}_pheno.tsv
cov=${sample}0${i}_cov.tsv
n=$(awk '{print NF; exit}' "$cov")

$plink2 \
    --bfile $sample \
    --pheno ${sample}0${i}_pheno.tsv \
    --maf 0.01 \
    --covar ${sample}0${i}_cov.tsv \
    --covar-col-nums 3-$n \
    --covar-variance-standardize \
    --glm \
    --out ${output}/${sample}0${i}
done; done
```
